Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-06-12 iaus/proceed.tex
minor update to iaus314

\section{The need for rapid cooling for the VSI}
Astrophysical disks generally have vertical shear, $\p_z\Omega\neq
0$. This is a source of free energy, and may thus lead to
instability (\cite[Goldreic \& Schubert 1967]{goldreich67};
\cite[Urpin 2003]{urpin03}; \cite[Barker \& Latter 2015]{barker15}),
provided the disturbance vertical lengthscale is much larger than its
radial lengthscale. However, if the disk is stably stratified, as is typical
for irradiated protoplanetary disks, the VSI also requires rapid
cooling to overcome the stabilizing influence of vertical buoyancy
(\cite[Nelson et al. 2013]{nelson13}).

More specifically, for a vertically isothermal disk with radial
temperature dependence $T\propto r^{q}$, the cooling or thermal
relaxation timescale $t_c$ must be sufficiently small for the VSI to
operate effectively:
\begin{align}\label{crit}
  t_c\Omega_\mathrm{K} < \frac{h|q|}{\gamma - 1}\equiv
  \beta_\mathrm{crit},  
\end{align}
where $\Omega_\mathrm{K}$ is the Keplerian frequency, $h$  is the disk
aspect ratio, and $\gamma$ is the adiabatic index (\cite[Lin \& Youdin
2015]{lin15}). 
Since protoplanetary disks are typically thin, $h\ll 1$,
Eq. \ref{crit} implies that $t_c \ll \Omega_\mathrm{K}$ is required
for the VSI. That is, the thermal timescale must be significantly
shorter than the dynamical timescale, which reflects the fact that
vertical shear is weakly destablizing, while vertical buoyancy is
strongly stabilizing.

@@

\section{Conclusions}
Vertical shear is ubiquitous in astrophysical disks. Such
configurations are unstable if the cooling timescales are
significantly shorter than the dynamical timescale (Eq. \ref{crit}). 
This can be satisified for small-scale disturbances at few 10s
of AU in typical protoplanetary disk 
models with realistic cooling. The VSI is thus dynamically important
in the outer disk, with potentially significantly implications for
transport and dust evolution.

>>>>

\section{The need for rapid cooling for the VSI}
Astrophysical disks generally have vertical shear, $\p_z\Omega\neq
0$. This is a source of free energy, and may thus lead to
instability (\cite[Goldreich \& Schubert 1967]{goldreich67};
\cite[Urpin 2003]{urpin03}; \cite[Barker \& Latter 2015]{barker15}),
provided the disturbance vertical lengthscale is much larger than its
radial lengthscale. However, if the disk is stably stratified, as is typical
for irradiated protoplanetary disks, the VSI also requires rapid
cooling to overcome the stabilizing influence of vertical buoyancy
(\cite[Nelson et al. 2013]{nelson13}).

More specifically, for a vertically isothermal disk with radial
temperature dependence $T\propto r^{q}$, the cooling or thermal
relaxation timescale $t_c$ must be sufficiently small for the VSI to
operate effectively:
\begin{align}\label{crit}
  t_c\Omega_\mathrm{K} < \frac{h|q|}{\gamma - 1}\equiv
  \beta_\mathrm{crit},  
\end{align}
where $\Omega_\mathrm{K}$ is the Keplerian frequency, $h$  is the disk
aspect ratio, and $\gamma$ is the adiabatic index (\cite[Lin \& Youdin
2015]{lin15}). 
Since protoplanetary disks are typically thin, $h\ll 1$,
Eq. \ref{crit} implies that $t_c \ll \Omega_\mathrm{K}$ is required
for the VSI. That is, the thermal timescale must be significantly
shorter than the dynamical timescale, which reflects the fact that
vertical shear is weakly destabilizing, while vertical buoyancy is
strongly stabilizing.

@@

\section{Conclusions}
Vertical shear is ubiquitous in astrophysical disks. Such
configurations are unstable if the cooling timescales are
significantly shorter than the dynamical timescale (Eq. \ref{crit}). 
This can be satisfied for small-scale disturbances at few 10s
of AU in typical protoplanetary disk 
models with realistic cooling. The VSI is thus dynamically important
in the outer disk, with potentially significantly implications for
transport and dust evolution.
